---
title: "Exploring Census PUMS Data with R"
author: "AMJ"
---

## Introduction

This project is based around developing functions with the Public Use Microdata (PUMS) Census API. The goal of this project is to learn how to work with R in a real-life application and to make something where users would be able to choose variables, survey years, and geographic subsets rather than rewriting the API request each time.

I first created the reusable functions, tested them, and did a small investigation.

## Setting Things Up

### Packages and API Key

These are all the packages I am using for this project. The `quantreg` and `ggplot2` packages are useful for creating plots!

```{r}
# Read packages
library(httr)
library(jsonlite)
library(tibble)
library(hms)
library(quantreg)
library(ggplot2)
```

#### Load the API Key

The API key is stored in a separate `.Renviron` file in the main project folder, which will allow the API key to be used without it being exposed in the GitHub commits. I did this by making a `.gitignore` file which doesn't push any files to GitHub with the `.Renviron` file type extension. The code below loads the API key.

```{r}
# Import API Key
readRenviron(".Renviron")
stopifnot(nzchar(Sys.getenv("CENSUS_API_KEY")))
```

## Data

### Inspecting the Variable Dictionary

I began with the given URL to understand the JSON structure before writing all the reusable functions. So, `GET()` downloads the response, `stop_for_status()` checks for HTTP errors, and `fromJSON()` parses the text into R lists. And setting `simplifyVector = FALSE` preserves the nested list structure.

The following example looks at `SEX`. Its dictionary includes a description and the mapping between codes and labels.

```{r}
url <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

response <- GET(url)
stop_for_status(response)

# Convert the response to text, then parse it into R lists
json_text <- content(response, as = "text", encoding = "UTF-8")
dictionary <- fromJSON(json_text, simplifyVector = FALSE)

# Inspect one variable
dictionary$variables$SEX
dictionary$variables$SEX$values
```

## Part 1: Data Processing

### Parsing Variable Values

I implemented that process in `get_variable_values(year)`. The function creates the URL for the selected year, keeps the required variables, and extracts each variable's `values`. Doing this keeps both category labels and numeric ranges.

One thing to note is that the state variable is named `ST` in 2021–2022 and `STATE` in 2023–2024, so the function is able to select the name according to the year.

```{r}
get_variable_values <- function(year) {

  # Build the URL for the chosen year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )

  response <- GET(url)
  stop_for_status(response)

  json_text <- content(response, as = "text", encoding = "UTF-8")
  dictionary <- fromJSON(json_text, simplifyVector = FALSE)

  # State variable name changes after 2022
  state_var <- if (year <= 2022) "ST" else "STATE"

  wanted <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_var, "REGION", "DIVISION"
  )

  # Keep only the requested variables
  selected <- dictionary$variables[wanted]

  # Extract each variable's values, preserving variable names
  result <- lapply(selected, function(variable) {
    variable$values
  })

  return(result)
}
```

Here, I'm calling the function for each year and saving the four named lists in one RDS file which is found in the data folder. All upcoming functions can read this file to interpret the returned codes using the appropriate year's dictionary.

```{r}
variable_values <- list(
  "2021" = get_variable_values(2021),
  "2022" = get_variable_values(2022),
  "2023" = get_variable_values(2023),
  "2024" = get_variable_values(2024)
)
```

#### Inspect the Selected Values

```{r}
# Check the result
names(variable_values[["2024"]])
variable_values[["2024"]]$SEX
```

#### Save the Dictionaries

```{r}
# Create the folder and save the four lists in one RDS file
dir.create("data", showWarnings = FALSE)
saveRDS(variable_values, file = "data/variable_values.rds")
```

### Parsing API Data

I divided the parsing work into many helper functions. Here is what they are and what they all do:

- `response_to_tibble()` takes a `GET()` response, uses its first row as column names, and converts the remaining rows into a tibble.
- `normalize_codes()` lets codes like `001` and `1` match.
- `time_midpoint()` converts a labeled time interval into its midpoint in seconds, which accounts for any intervals crossing midnight as well as A.M. and P.M.
- `parse_time_codes()` looks up time labels and returns `hms` time-of-day values. Missing or unparseable time intervals become `NA`.
- `parse_category()` converts category codes into labeled factors.

The main `query_pums()` function validates the year, variable selections, geography level, and geography codes before calling the API. It requires `PWGTP`, at least one other numeric variable, and at least one categorical variable. Defaults select age and sex in North Carolina in 2024. Region and Division have separate default codes, and `geography_values = "all"` requests all values within the selected geography level.

After downloading, the function converts columns to their appropriate types and adds the `census` class. This class is what enables the custom summary and plot methods below.

#### Convert an API Response into a Tibble

```{r}
response_to_tibble <- function(response) {
  stop_for_status(response)

  text <- content(response, as = "text", encoding = "UTF-8")

  if (!jsonlite::validate(text)) {
    stop("The API did not return JSON. Check your query and API key.")
  }

  rows <- fromJSON(text, simplifyVector = FALSE)

  if (length(rows) == 0) {
    stop("The API returned no data.")
  }

  headers <- unlist(rows[[1]], use.names = FALSE)

  columns <- lapply(seq_along(headers), function(j) {
    vapply(rows[-1], function(row) {
      if (is.null(row[[j]])) NA_character_ else as.character(row[[j]])
    }, character(1))
  })

  names(columns) <- headers
  as_tibble(columns)
}
```

#### Normalize Variable Codes

```{r}
# Make codes such as "001" and "1" match
normalize_codes <- function(x) {
  x <- trimws(as.character(x))
  numeric_codes <- !is.na(x) & grepl("^[0-9]+$", x)
  x[numeric_codes] <- as.character(as.integer(x[numeric_codes]))
  x
}
```

#### Find the Midpoint of a Time Interval

```{r}
# Convert one time interval label to its midpoint in seconds
time_midpoint <- function(label) {
  if (is.na(label)) {
    return(NA_real_)
  }

  label <- tolower(gsub("\\.", "", label))
  endpoints <- strsplit(label, "\\s+to\\s+")[[1]]

  if (length(endpoints) != 2) {
    return(NA_real_)
  }

  clock_seconds <- function(x) {
    pattern <- "^\\s*([0-9]{1,2}):([0-9]{2})\\s*(am|pm)\\s*$"
    parts <- regmatches(x, regexec(pattern, x))[[1]]

    if (length(parts) != 4) {
      return(NA_real_)
    }

    hour <- as.numeric(parts[2]) %% 12
    minute <- as.numeric(parts[3])

    if (parts[4] == "pm") {
      hour <- hour + 12
    }

    hour * 3600 + minute * 60
  }

  start <- clock_seconds(endpoints[1])
  end <- clock_seconds(endpoints[2])

  if (is.na(start) || is.na(end)) {
    return(NA_real_)
  }

  # Handle intervals that cross midnight
  if (end < start) {
    end <- end + 86400
  }

  ((start + end) / 2) %% 86400
}
```

#### Convert Time Codes into Time Values

```{r}
# Look up time codes and return time-of-day values
parse_time_codes <- function(x, variable_values) {
  labels <- unlist(variable_values$item, use.names = TRUE)
  midpoints <- vapply(labels, time_midpoint, numeric(1))

  index <- match(
    normalize_codes(x),
    normalize_codes(names(labels))
  )

  hms::as_hms(unname(midpoints[index]))
}
```

#### Convert Categorical Codes into Factors

```{r}
# Convert categorical codes into labeled factors
parse_category <- function(x, variable_values) {
  labels <- unlist(variable_values$item, use.names = TRUE)
  codes <- normalize_codes(names(labels))

  # Sort levels by their underlying numeric codes
  ordering <- order(suppressWarnings(as.numeric(codes)), codes)
  labels <- labels[ordering]
  codes <- codes[ordering]

  factor(
    normalize_codes(x),
    levels = codes,
    labels = unname(labels)
  )
}
```

#### Query and Clean One Survey Year

```{r}
query_pums <- function(
    year = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_values = NULL,
    rds_file = "data/variable_values.rds",
    api_key = Sys.getenv("CENSUS_API_KEY", unset = "")
) {
  # Validate year
  if (length(year) != 1 || is.na(year) ||
      !(year %in% 2021:2024)) {
    stop("year must be one of 2021, 2022, 2023, or 2024.")
  }

  allowed_numeric <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
  )

  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")

  # Validate numeric variables
  if (!is.character(numeric_vars) || anyNA(numeric_vars) ||
      !all(numeric_vars %in% allowed_numeric)) {
    stop("Invalid numeric variable.")
  }

  if (!("PWGTP" %in% numeric_vars) ||
      length(setdiff(numeric_vars, "PWGTP")) == 0) {
    stop("Include PWGTP and at least one other numeric variable.")
  }

  # Validate categorical variables
  if (!is.character(categorical_vars) ||
      length(categorical_vars) == 0 ||
      anyNA(categorical_vars) ||
      !all(categorical_vars %in% allowed_categorical)) {
    stop("Include at least one valid categorical variable.")
  }

  # Validate geography level
  if (!is.character(geography) || length(geography) != 1 ||
      is.na(geography) ||
      !(tolower(geography) %in% c("region", "division", "state"))) {
    stop("geography must be exactly one of Region, Division, or State.")
  }

  geography <- tolower(geography)
  numeric_vars <- unique(numeric_vars)
  categorical_vars <- unique(categorical_vars)

  # Load the dictionaries saved in Part 1
  dictionaries <- readRDS(rds_file)
  metadata <- dictionaries[[as.character(year)]]

  if (is.null(metadata)) {
    stop("The RDS file does not contain the requested year.")
  }

  geography_variable <- switch(
    geography,
    state = if (year <= 2022) "ST" else "STATE",
    region = "REGION",
    division = "DIVISION"
  )

  geography_labels <- unlist(
    metadata[[geography_variable]]$item,
    use.names = TRUE
  )

  valid_codes <- normalize_codes(names(geography_labels))

  # Defaults: North Carolina, South Atlantic, or South
  if (is.null(geography_values)) {
    geography_values <- switch(
      geography,
      state = "37",
      division = "5",
      region = "3"
    )
  }

  if (length(geography_values) == 0 || anyNA(geography_values)) {
    stop("Provide valid geography codes or 'all'.")
  }

  geography_values <- as.character(geography_values)

  if (any(tolower(geography_values) == "all")) {
    if (length(geography_values) != 1) {
      stop("'all' cannot be combined with specific geography codes.")
    }

    geography_filter <- "*"
  } else {
    codes <- normalize_codes(geography_values)

    if (!all(codes %in% valid_codes)) {
      stop("Invalid geography code. Check the selected year's RDS values.")
    }

    # State FIPS codes must have two digits
    if (geography == "state") {
      codes <- sprintf("%02d", as.integer(codes))
    }

    geography_filter <- paste(unique(codes), collapse = ",")
  }

  # Build and submit the API query
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums"
  )

  query <- list(
    get = paste(c(numeric_vars, categorical_vars), collapse = ",")
  )

  query[["for"]] <- paste0(geography, ":", geography_filter)

  if (nzchar(api_key)) {
    query$key <- api_key
  }

  response <- GET(url, query = query)
  census_data <- response_to_tibble(response)

  # Convert numeric and time variables
  for (variable in numeric_vars) {
    if (variable %in% c("JWAP", "JWDP")) {
      census_data[[variable]] <- parse_time_codes(
        census_data[[variable]],
        metadata[[variable]]
      )
    } else {
      census_data[[variable]] <- suppressWarnings(
        as.numeric(census_data[[variable]])
      )
    }
  }

  # Convert categorical variables using the correct year's labels
  for (variable in categorical_vars) {
    census_data[[variable]] <- parse_category(
      census_data[[variable]],
      metadata[[variable]]
    )
  }

  class(census_data) <- c("census", class(census_data))
  return(census_data)
}
```

### Combining Years

`query_pums_years()` calls the single-year function for each requested year. Each result receives a `year` column, and `bind_rows()` combines the results into one census tibble. Assigning the returned value to an object like `combined_data` makes it available to our later functions and function calls.

```{r}
query_pums_years <- function(
    years = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_values = NULL,
    rds_file = "data/variable_values.rds",
    api_key = Sys.getenv("CENSUS_API_KEY", unset = "")
) {
  if (length(years) == 0 || anyNA(years) ||
      !all(years %in% 2021:2024)) {
    stop("years must contain only 2021, 2022, 2023, or 2024.")
  }

  yearly_data <- lapply(unique(years), function(year) {
    data <- query_pums(
      year = year,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geography_values = geography_values,
      rds_file = rds_file,
      api_key = api_key
    )

    data$year <- as.integer(year)
    data
  })

  combined_data <- dplyr::bind_rows(yearly_data)

  class(combined_data) <- unique(c("census", class(combined_data)))
  return(combined_data)
}
```

## Part 2: Writing a Generic Function for Summarizing

### Generic Functions

R uses the `census` class to run our custom `summary.census()` and `plot.census()` functions when we call `summary()` or `plot()`.

The summary function returns a named list containing:

- Weighted means and standard deviations for numeric variables.
- Weighted counts for categorical variables.

By default, it summarizes all eligible columns. It excludes `PWGTP` and time variables from numeric summaries and treats `year` as a categorical variable. Users can choose specific columns or use `character(0)` to skip either type of summary.

Here, $x_i$ is a numeric value and $w_i$ is its respective `PWGTP` weight. Missing numeric values and their weights are not included in that calculation. However, missing categorical values are counted separately.

The plotting function creates a boxplot for one numeric variable across the groups of one categorical variable. It includes `weight = PWGTP` to account for each row's weight.

```{r}
# Weighted summaries: called automatically by summary(census_data)
summary.census <- function(
    object,
    numeric_vars = NULL,
    categorical_vars = NULL
) {
  # Find numeric columns, excluding time variables, weights, and year
  numeric_columns <- names(object)[
    vapply(object, function(x) {
      is.numeric(x) &&
        !inherits(x, c("difftime", "POSIXt", "Date"))
    }, logical(1))
  ]

  numeric_columns <- setdiff(numeric_columns, c("PWGTP", "year"))

  # Find categorical columns, including year
  categorical_columns <- names(object)[
    vapply(object, function(x) {
      is.factor(x) || is.character(x) || is.logical(x)
    }, logical(1))
  ]

  categorical_columns <- union(
    categorical_columns,
    intersect("year", names(object))
  )

  # Default to all eligible columns
  if (is.null(numeric_vars)) {
    numeric_vars <- numeric_columns
  }

  if (is.null(categorical_vars)) {
    categorical_vars <- categorical_columns
  }

  # Check the selected variables
  if (!is.character(numeric_vars) || anyNA(numeric_vars) ||
      !all(numeric_vars %in% numeric_columns)) {
    stop("Choose valid numeric variables, excluding PWGTP, year, and times.")
  }

  if (!is.character(categorical_vars) || anyNA(categorical_vars) ||
      !all(categorical_vars %in% categorical_columns)) {
    stop("Choose valid categorical variables or year.")
  }

  weights <- object$PWGTP

  if (!is.numeric(weights) || any(!is.finite(weights)) ||
      any(weights < 0)) {
    stop("PWGTP must contain finite, nonnegative numeric weights.")
  }

  # Calculate weighted means and standard deviations
  numeric_results <- list()

  for (variable in numeric_vars) {
    numeric_vector <- object[[variable]]
    weight_vector <- weights

    # Remove missing values and their corresponding weights
    keep <- is.finite(numeric_vector) & weight_vector > 0
    numeric_vector <- numeric_vector[keep]
    weight_vector <- weight_vector[keep]

    if (length(numeric_vector) == 0) {
      sample_mean <- NA_real_
      sample_sd <- NA_real_
    } else {
      sample_mean <- sum(numeric_vector * weight_vector) /
        sum(weight_vector)

      sample_sd <- sqrt(
        sum(numeric_vector^2 * weight_vector) /
          sum(weight_vector) - sample_mean^2
      )
    }

    numeric_results[[variable]] <- c(
      mean = sample_mean,
      sd = sample_sd
    )
  }

  # Calculate weighted counts for categorical variables
  categorical_results <- list()

  for (variable in categorical_vars) {
    category <- object[[variable]]

    if (!is.factor(category)) {
      category <- factor(category)
    }

    category <- addNA(category, ifany = TRUE)

    counts <- tapply(weights, category, sum, default = 0)

    categorical_results[[variable]] <- setNames(
      as.numeric(counts),
      names(counts)
    )
  }

  # Return a named list
  return(list(
    numeric = numeric_results,
    categorical = categorical_results
  ))
}
```

### Graphical Summaries

This method creates a weighted boxplot for the chosen numeric and categorical columns.

```{r}
# Called automatically by plot(census_data, ...)
plot.census <- function(x, cat_var, num_var, ...) {
  if (!is.character(cat_var) || length(cat_var) != 1 ||
      is.na(cat_var) || !(cat_var %in% names(x))) {
    stop("Specify one categorical column name as a string.")
  }

  if (!is.character(num_var) || length(num_var) != 1 ||
      is.na(num_var) || !(num_var %in% names(x))) {
    stop("Specify one numeric column name as a string.")
  }

  category <- x[[cat_var]]
  numeric_value <- x[[num_var]]

  if (!(cat_var == "year" || is.factor(category) ||
        is.character(category) || is.logical(category))) {
    stop("cat_var must be categorical or year.")
  }

  if (!is.numeric(numeric_value) ||
      inherits(numeric_value, c("difftime", "POSIXt", "Date")) ||
      num_var %in% c("PWGTP", "year")) {
    stop("num_var must be numeric, excluding PWGTP, year, and times.")
  }

  if (!is.numeric(x$PWGTP) || any(!is.finite(x$PWGTP)) ||
      any(x$PWGTP < 0)) {
    stop("PWGTP must contain finite, nonnegative numeric weights.")
  }

  keep <- !is.na(category) & is.finite(numeric_value) & x$PWGTP > 0
  plot_data <- x[keep, ]

  if (nrow(plot_data) == 0) {
    stop("No complete observations with positive weights to plot.")
  }

  plot_data[[cat_var]] <- factor(plot_data[[cat_var]])

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data[[cat_var]],
      y = .data[[num_var]],
      weight = .data[["PWGTP"]]
    )
  ) +
    ggplot2::geom_boxplot() +
    ggplot2::labs(
      x = cat_var,
      y = num_var,
      title = paste(num_var, "by", cat_var)
    ) +
    ggplot2::theme_minimal()
}
```

## Testing the Functions

### Dictionary structure and time parsing

These checks verify that all four dictionaries contain the required variables and that time conversion handles morning, noon, midnight, and missing values. A success message appears only if checks pass.

```{r}
for (y in 2021:2024) {
  state_name <- if (y <= 2022) "ST" else "STATE"
  expected <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                state_name, "REGION", "DIVISION")
  d <- variable_values[[as.character(y)]]
  stopifnot(setequal(names(d), expected),
            all(vapply(d, function(x) !is.null(x), logical(1))))
}
```

#### Test Time Midpoints

```{r}
stopifnot(
  time_midpoint("6:00 a.m. to 6:04 a.m.") == 21720,
  time_midpoint("12:00 p.m. to 12:04 p.m.") == 43320,
  time_midpoint("12:00 a.m. to 12:04 a.m.") == 120,
  is.na(time_midpoint("N/A (not a worker)"))
)
cat("Dictionary and time parsing checks passed.")
```

### Checking weighted calculations by hand

For ages 20 and 40, which have weights 1 and 3, the weighted mean is 35 and the weighted standard deviation is $\sqrt{75}$, or about 8.66. The weighted counts should be 1 for group A and 3 for group B. This example checks the math independently of the API.

```{r}
toy_data <- tibble::tibble(
  AGEP = c(20, 40), PWGTP = c(1, 3), group = factor(c("A", "B"))
)
class(toy_data) <- c("census", class(toy_data))
toy_summary <- summary(toy_data)
stopifnot(
  isTRUE(all.equal(unname(toy_summary$numeric$AGEP["mean"]), 35)),
  isTRUE(all.equal(unname(toy_summary$numeric$AGEP["sd"]), sqrt(75))),
  toy_summary$categorical$group["A"] == 1,
  toy_summary$categorical$group["B"] == 3
)
toy_summary
cat("Weighted calculation checks passed.")
```

### Single-year and multiple-year downloads

I tested the defaults with North Carolina and the multiple year function with Wyoming (which has the state code of `56`). 
These are separate datasets.. `nc_data` describes North Carolina in 2024 and `combined_data` describes Wyoming in 2021–2024.

```{r}
combined_data <- query_pums_years(
  years = 2021:2024,
  geography = "State",
  geography_values = "56"  
)
```

#### Download One Year

```{r}
file.exists("data/variable_values.rds")
nc_data <- query_pums()
```

```{r}
stopifnot(
  nrow(nc_data) > 0,
  inherits(nc_data, "census"),
  is.numeric(nc_data$AGEP), is.numeric(nc_data$PWGTP),
  is.factor(nc_data$SEX),
  inherits(combined_data, "census"),
  setequal(unique(combined_data$year), 2021:2024)
)
knitr::kable(as.data.frame(table(combined_data$year)),
             col.names = c("Survey year", "Downloaded records"))
cat("Single-year and multiple-year checks passed.")
```

### Summary and plot methods with different inputs

First, I called the default summary and then selected age and sex. I also selected `year` in the combined data. The plots use `sex` for North Carolina and `year` for Wyoming.

The combined summary combines the four yearly samples.

```{r}
# Summarize all eligible variables
summary(nc_data)
```

#### Summarize Selected Columns

```{r}
# Summarize selected variables
summary(
  nc_data,
  numeric_vars = "AGEP",
  categorical_vars = "SEX"
)
```

#### Plot Age by Sex

```{r}
# Weighted age boxplots by sex
plot(nc_data, cat_var = "SEX", num_var = "AGEP")+ggplot2::theme_gray()
```

#### Summarize Multiple Years

```{r}
# Summarize and plot multiple years
summary(
  combined_data,
  numeric_vars = "AGEP",
  categorical_vars = c("SEX", "year")
)
```

#### Plot Age by Year

```{r}
plot(combined_data, cat_var = "year", num_var = "AGEP")+ggplot2::theme_gray()
```

### Invalid inputs

The following tests provide invalid arguments. Each test checks that an error occurs and displays the error message. This is just to check if the functions reject invalid requests/data properly.

```{r}
expect_error <- function(expr) {
  message <- tryCatch({ force(expr); NULL },
                      error = function(e) conditionMessage(e))
  if (is.null(message)) stop("Test failed: an error was expected.")
  message
}
```

#### Check Invalid Requests

```{r}
error_checks <- list(
  "Invalid year" = expect_error(query_pums(year = 2020)),
  "Multiple years in single-year query" = expect_error(query_pums(year = c(2023, 2024))),
  "Missing weight" = expect_error(query_pums(numeric_vars = "AGEP")),
  "Only weight" = expect_error(query_pums(numeric_vars = "PWGTP")),
  "Invalid numeric variable" = expect_error(query_pums(numeric_vars = c("PWGTP", "BAD"))),
  "Missing category" = expect_error(query_pums(categorical_vars = character(0))),
  "Invalid category" = expect_error(query_pums(categorical_vars = "BAD")),
  "Multiple geography levels" = expect_error(query_pums(geography = c("State", "Region"))),
  "Invalid geography level" = expect_error(query_pums(geography = "All")),
  "Invalid state code" = expect_error(query_pums(geography_values = "99")),
  "Invalid plot input" = expect_error(plot(nc_data, cat_var = "AGEP", num_var = "SEX"))
)
knitr::kable(data.frame(Test = names(error_checks),
                        Message = unlist(error_checks, use.names = FALSE)))
```


## Investigation

### School Enrollment Among Young Adults in North Carolina

#### Question and Data Selection

How does school enrollment vary by age group and sex among North Carolina residents aged 18–29 in 2024?

I used age (`AGEP`), school enrollment (`SCH`), sex (`SEX`), and person weights (`PWGTP`). School enrollment includes school, college, and home school, so it doesn't mean specifically college enrollment alone. The Census Variable dictionary also identifies people who attended public or private school from those who didn't attend either in the data.

```{r}
education_data <- query_pums(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP"),
  categorical_vars = c("SCH", "SEX"),
  geography = "State",
  geography_values = "37"
)
```

#### Select Young Adults

To start off, I read the school enrollment labels from the dictionary. 

Code 1 indicates no enrollment while codes 2 and 3 indicate enrollment. I also made sure to exclude records outside ages 18–29, records without a valid enrollment or sex category, and records without a positive weight. This is because I'm looking at the ages most likely to go to school.

```{r}
dictionary_2024 <- readRDS("data/variable_values.rds")[["2024"]]
school_labels <- unlist(dictionary_2024$SCH$item, use.names = TRUE)
school_codes <- normalize_codes(names(school_labels))

enrolled_labels <- unname(school_labels[school_codes %in% c("2", "3")])
not_enrolled_label <- unname(school_labels[school_codes == "1"])

stopifnot(length(enrolled_labels) == 2,
          length(not_enrolled_label) == 1)

keep <- with(
  education_data,
  is.finite(AGEP) & AGEP >= 18 & AGEP <= 29 &
    is.finite(PWGTP) & PWGTP > 0 &
    !is.na(SEX) &
    SCH %in% c(enrolled_labels, not_enrolled_label)
)

young_adults <- education_data[keep, ]
stopifnot(nrow(young_adults) > 0,
          inherits(young_adults, "census"))
```

#### Create Enrollment and Age Groups

I combined public and private enrollment into one category and divided the age range I created into three different groups of four years each to make some categories.

```{r}
young_adults$enrollment <- factor(
  ifelse(young_adults$SCH %in% enrolled_labels,
         "Enrolled", "Not enrolled"),
  levels = c("Not enrolled", "Enrolled")
)

young_adults$age_group <- cut(
  young_adults$AGEP,
  breaks = c(18, 22, 26, 30),
  right = FALSE,
  labels = c("18–21", "22–25", "26–29")
)
```

#### Overall Weighted Summary

I used my summary method to calculate the mean age, the standard deviation of age, and the weighted categorical counts.

```{r}
education_summary <- summary(
  young_adults,
  numeric_vars = "AGEP",
  categorical_vars = c("enrollment", "age_group", "SEX")
)
education_summary
```

```{r}
enrollment_counts <- education_summary$categorical$enrollment
overall_enrollment_pct <- 100 * enrollment_counts["Enrolled"] /
  sum(enrollment_counts)

knitr::kable(
  data.frame(
    Sample_records = nrow(young_adults),
    Weighted_population = sum(young_adults$PWGTP),
    Mean_age = unname(education_summary$numeric$AGEP["mean"]),
    Enrollment_percent = unname(overall_enrollment_pct)
  ),
  digits = 1,
  col.names = c("Sample records", "Estimated people",
                "Mean age", "Enrolled (%)")
)
```

Among the included young adults, about **33.33%** were enrolled. The weighted mean age was **23.4 years**.

#### Compare Enrollment Percentages

Raw enrollment counts depend on how many people are in each group. To compare groups, I divided each group's weighted enrolled count by its total weighted count.

This helper function calls my summary method separately for each of the age groups. It also demonstrates a categorical-only summary by supplying `character(0)` for numeric variables.

```{r}
enrollment_by_group <- function(data, group_var) {
  groups <- levels(droplevels(factor(data[[group_var]])))

  rows <- lapply(groups, function(group) {
    group_data <- data[as.character(data[[group_var]]) == group, ]

    group_summary <- summary(
      group_data,
      numeric_vars = character(0),
      categorical_vars = "enrollment"
    )

    counts <- group_summary$categorical$enrollment

    tibble::tibble(
      group = group,
      sample_records = nrow(group_data),
      weighted_population = sum(counts),
      weighted_enrolled = unname(counts["Enrolled"]),
      enrollment_percent = 100 * unname(counts["Enrolled"]) / sum(counts)
    )
  })

  dplyr::bind_rows(rows)
}
```

#### Enrollment by Age Group

```{r}
enrollment_by_age <- enrollment_by_group(young_adults, "age_group")

knitr::kable(
  enrollment_by_age,
  digits = 1,
  col.names = c("Age group", "Sample records", "Estimated people",
                "Estimated enrolled", "Enrolled (%)")
)
```

The percentages in this table compare enrollment within each age group, rather than each group's share of all enrolled people. The age group `18-21` had the highest enrollment percentage with **62.2%**, while the age group `26-29` had the lowest enrollment percentage with **12.6%**.

```{r}
ggplot2::ggplot(
  enrollment_by_age,
  ggplot2::aes(x = group, y = enrollment_percent)
) +
  ggplot2::geom_col(fill = "steelblue") +
  ggplot2::scale_y_continuous(limits = c(0, 100)) +
  ggplot2::labs(
    title = "School enrollment by age group",
    subtitle = "North Carolina residents aged 18–29, 2024",
    x = "Age group",
    y = "Enrolled (%)"
  ) +
  ggplot2::theme_gray()
```

This chart shows us the difference between the age groups and enrollment percentage! The 18-21 category dwarfs the other categories by a good amount. It makes sense though as that is when most students graduate high school and are going through college.

#### Enrollment by Sex

```{r}
enrollment_by_sex <- enrollment_by_group(young_adults, "SEX")

knitr::kable(
  enrollment_by_sex,
  digits = 1,
  col.names = c("Sex", "Sample records", "Estimated people",
                "Estimated enrolled", "Enrolled (%)")
)
```

The sex comparison uses the categories supplied by the Census. We can see that a larger percentage of women are estimated to be enrolled.

#### Weighted Boxplot: Age by Enrollment

I use my plotting method to compare the age distributions of enrolled and non-enrolled young adults. The boxplots show age distributions, while the tables above show enrollment percentages.

```{r}
plot(young_adults, cat_var = "enrollment", num_var = "AGEP") +
  ggplot2::labs(
    title = "Age by school enrollment",
    subtitle = "North Carolina residents aged 18–29, 2024",
    x = "School enrollment",
    y = "Age (years)"
  ) +
  ggplot2::theme_gray()
```

#### Weighted Boxplot: Age by Sex Among Enrolled Young Adults

For a second use of my plotting method, I selected enrolled young adults and change the categorical variable to sex.

```{r}
enrolled_adults <- young_adults[young_adults$enrollment == "Enrolled", ]
stopifnot(nrow(enrolled_adults) > 0)

plot(enrolled_adults, cat_var = "SEX", num_var = "AGEP") +
  ggplot2::labs(
    title = "Age by sex among enrolled young adults",
    subtitle = "North Carolina residents aged 18–29, 2024",
    x = "Sex",
    y = "Age (years)"
  ) +
  ggplot2::theme_gray()
```

You can see that when comparing sex, the boxplots are quite similar in terms of the minimum, Q1, median, Q3, and maximum. This means that there aren't many differences between the age distributions of men vs. women.

Next, I used my summary method with only a numeric variable to describe the ages of enrolled young adults.

```{r}
enrolled_age_summary <- summary(
  enrolled_adults,
  numeric_vars = "AGEP",
  categorical_vars = character(0)
)
enrolled_age_summary
```

#### Conclusions

The highest estimated enrollment percentage was in the **18-21** age group(s). The difference between the highest and lowest age-group percentages was **49.6 percentage points**. Across the sex categories, the difference was **5.7 percentage points**. Among enrolled young adults, the weighted mean age was **23.4 years**.

These results use data from subjects from North Carolina aged 18-29 within the 2024 sample. The results haven't been investigated further using statistical analysis or anything to make sure that they are actually significant, so the results will have to be taken with a grain of salt. But otherwise, I had a lot of fun doing this project and I learned a lot about the power of reusable functions within R.
