---
title: "Exploring Census PUMS Data with R"
author: "AMJ"
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
    theme:
      light: minty
      dark: vapor
editor: visual
---

## Introduction

This project is based around developing functions with the Public Use Microdata (PUMS) Census API. The goal of this project is to learn how to work with R in a real-life application and to make something where users would be able to choose variables, survey years, and geographic subsets rather than rewriting the API request each time.

## Setup

The API key is stored in a separate `.Renviron` file in the main project folder, which will allow the API key to be used without it being exposed in the GitHub commits. I did this by making a `.gitignore` file which doesn't push any files to GitHub with the `.Renviron` file type extension.

The code below loads the key and also all the packages that I will need to complete this project. The `quantreg` and `ggplot2` packages are useful for creating plots!

```{r}
library(httr)
library(jsonlite)
library(tibble)
library(hms)
library(quantreg)
library(ggplot2)
readRenviron(".Renviron")
stopifnot(nzchar(Sys.getenv("CENSUS_API_KEY")))
```

## Creating the functions

```{r}
url <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

response <- GET(url)
stop_for_status(response)

# Convert the response to text, then parse it into R lists
json_text <- content(response, as = "text", encoding = "UTF-8")
dictionary <- fromJSON(json_text, simplifyVector = FALSE)

# Inspect one variable
dictionary$variables$SEX
dictionary$variables$SEX$values
```

```{r}
get_variable_values <- function(year) {

  # Build the URL for the chosen year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )

  response <- GET(url)
  stop_for_status(response)

  json_text <- content(response, as = "text", encoding = "UTF-8")
  dictionary <- fromJSON(json_text, simplifyVector = FALSE)

  # State variable name changes after 2022
  state_var <- if (year <= 2022) "ST" else "STATE"

  wanted <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_var, "REGION", "DIVISION"
  )

  # Keep only the requested variables
  selected <- dictionary$variables[wanted]

  # Extract each variable's values, preserving variable names
  result <- lapply(selected, function(variable) {
    variable$values
  })

  return(result)
}
```

```{r}
variable_values <- list(
  "2021" = get_variable_values(2021),
  "2022" = get_variable_values(2022),
  "2023" = get_variable_values(2023),
  "2024" = get_variable_values(2024)
)

# Check the result
names(variable_values[["2024"]])
variable_values[["2024"]]$SEX

# Create the folder and save the four lists in one RDS file
dir.create("data", showWarnings = FALSE)

saveRDS(variable_values, file = "data/variable_values.rds")
```

```{r}
response_to_tibble <- function(response) {
  stop_for_status(response)

  text <- content(response, as = "text", encoding = "UTF-8")

  if (!jsonlite::validate(text)) {
    stop("The API did not return JSON. Check your query and API key.")
  }

  rows <- fromJSON(text, simplifyVector = FALSE)

  if (length(rows) == 0) {
    stop("The API returned no data.")
  }

  headers <- unlist(rows[[1]], use.names = FALSE)

  columns <- lapply(seq_along(headers), function(j) {
    vapply(rows[-1], function(row) {
      if (is.null(row[[j]])) NA_character_ else as.character(row[[j]])
    }, character(1))
  })

  names(columns) <- headers
  as_tibble(columns)
}

# Make codes such as "001" and "1" match
normalize_codes <- function(x) {
  x <- trimws(as.character(x))
  numeric_codes <- !is.na(x) & grepl("^[0-9]+$", x)
  x[numeric_codes] <- as.character(as.integer(x[numeric_codes]))
  x
}

# Convert one time interval label to its midpoint in seconds
time_midpoint <- function(label) {
  if (is.na(label)) {
    return(NA_real_)
  }

  label <- tolower(gsub("\\.", "", label))
  endpoints <- strsplit(label, "\\s+to\\s+")[[1]]

  if (length(endpoints) != 2) {
    return(NA_real_)
  }

  clock_seconds <- function(x) {
    pattern <- "^\\s*([0-9]{1,2}):([0-9]{2})\\s*(am|pm)\\s*$"
    parts <- regmatches(x, regexec(pattern, x))[[1]]

    if (length(parts) != 4) {
      return(NA_real_)
    }

    hour <- as.numeric(parts[2]) %% 12
    minute <- as.numeric(parts[3])

    if (parts[4] == "pm") {
      hour <- hour + 12
    }

    hour * 3600 + minute * 60
  }

  start <- clock_seconds(endpoints[1])
  end <- clock_seconds(endpoints[2])

  if (is.na(start) || is.na(end)) {
    return(NA_real_)
  }

  # Handle intervals that cross midnight
  if (end < start) {
    end <- end + 86400
  }

  ((start + end) / 2) %% 86400
}

# Look up time codes and return time-of-day values
parse_time_codes <- function(x, variable_values) {
  labels <- unlist(variable_values$item, use.names = TRUE)
  midpoints <- vapply(labels, time_midpoint, numeric(1))

  index <- match(
    normalize_codes(x),
    normalize_codes(names(labels))
  )

  hms::as_hms(unname(midpoints[index]))
}

# Convert categorical codes into labeled factors
parse_category <- function(x, variable_values) {
  labels <- unlist(variable_values$item, use.names = TRUE)
  codes <- normalize_codes(names(labels))

  # Sort levels by their underlying numeric codes
  ordering <- order(suppressWarnings(as.numeric(codes)), codes)
  labels <- labels[ordering]
  codes <- codes[ordering]

  factor(
    normalize_codes(x),
    levels = codes,
    labels = unname(labels)
  )
}

query_pums <- function(
    year = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_values = NULL,
    rds_file = "data/variable_values.rds",
    api_key = Sys.getenv("CENSUS_API_KEY", unset = "")
) {
  # Validate year
  if (length(year) != 1 || is.na(year) ||
      !(year %in% 2021:2024)) {
    stop("year must be one of 2021, 2022, 2023, or 2024.")
  }

  allowed_numeric <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
  )

  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")

  # Validate numeric variables
  if (!is.character(numeric_vars) || anyNA(numeric_vars) ||
      !all(numeric_vars %in% allowed_numeric)) {
    stop("Invalid numeric variable.")
  }

  if (!("PWGTP" %in% numeric_vars) ||
      length(setdiff(numeric_vars, "PWGTP")) == 0) {
    stop("Include PWGTP and at least one other numeric variable.")
  }

  # Validate categorical variables
  if (!is.character(categorical_vars) ||
      length(categorical_vars) == 0 ||
      anyNA(categorical_vars) ||
      !all(categorical_vars %in% allowed_categorical)) {
    stop("Include at least one valid categorical variable.")
  }

  # Validate geography level
  if (!is.character(geography) || length(geography) != 1 ||
      is.na(geography) ||
      !(tolower(geography) %in% c("region", "division", "state"))) {
    stop("geography must be exactly one of Region, Division, or State.")
  }

  geography <- tolower(geography)
  numeric_vars <- unique(numeric_vars)
  categorical_vars <- unique(categorical_vars)

  # Load the dictionaries saved in Part 1
  dictionaries <- readRDS(rds_file)
  metadata <- dictionaries[[as.character(year)]]

  if (is.null(metadata)) {
    stop("The RDS file does not contain the requested year.")
  }

  geography_variable <- switch(
    geography,
    state = if (year <= 2022) "ST" else "STATE",
    region = "REGION",
    division = "DIVISION"
  )

  geography_labels <- unlist(
    metadata[[geography_variable]]$item,
    use.names = TRUE
  )

  valid_codes <- normalize_codes(names(geography_labels))

  # Defaults: North Carolina, South Atlantic, or South
  if (is.null(geography_values)) {
    geography_values <- switch(
      geography,
      state = "37",
      division = "5",
      region = "3"
    )
  }

  if (length(geography_values) == 0 || anyNA(geography_values)) {
    stop("Provide valid geography codes or 'all'.")
  }

  geography_values <- as.character(geography_values)

  if (any(tolower(geography_values) == "all")) {
    if (length(geography_values) != 1) {
      stop("'all' cannot be combined with specific geography codes.")
    }

    geography_filter <- "*"
  } else {
    codes <- normalize_codes(geography_values)

    if (!all(codes %in% valid_codes)) {
      stop("Invalid geography code. Check the selected year's RDS values.")
    }

    # State FIPS codes must have two digits
    if (geography == "state") {
      codes <- sprintf("%02d", as.integer(codes))
    }

    geography_filter <- paste(unique(codes), collapse = ",")
  }

  # Build and submit the API query
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums"
  )

  query <- list(
    get = paste(c(numeric_vars, categorical_vars), collapse = ",")
  )

  query[["for"]] <- paste0(geography, ":", geography_filter)

  if (nzchar(api_key)) {
    query$key <- api_key
  }

  response <- GET(url, query = query)
  census_data <- response_to_tibble(response)

  # Convert numeric and time variables
  for (variable in numeric_vars) {
    if (variable %in% c("JWAP", "JWDP")) {
      census_data[[variable]] <- parse_time_codes(
        census_data[[variable]],
        metadata[[variable]]
      )
    } else {
      census_data[[variable]] <- suppressWarnings(
        as.numeric(census_data[[variable]])
      )
    }
  }

  # Convert categorical variables using the correct year's labels
  for (variable in categorical_vars) {
    census_data[[variable]] <- parse_category(
      census_data[[variable]],
      metadata[[variable]]
    )
  }

  class(census_data) <- c("census", class(census_data))
  return(census_data)
}
```

```{r}
query_pums_years <- function(
    years = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_values = NULL,
    rds_file = "data/variable_values.rds",
    api_key = Sys.getenv("CENSUS_API_KEY", unset = "")
) {
  if (length(years) == 0 || anyNA(years) ||
      !all(years %in% 2021:2024)) {
    stop("years must contain only 2021, 2022, 2023, or 2024.")
  }

  yearly_data <- lapply(unique(years), function(year) {
    data <- query_pums(
      year = year,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geography_values = geography_values,
      rds_file = rds_file,
      api_key = api_key
    )

    data$year <- as.integer(year)
    data
  })

  combined_data <- dplyr::bind_rows(yearly_data)

  class(combined_data) <- unique(c("census", class(combined_data)))
  return(combined_data)
}
```

```{r}
combined_data <- query_pums_years(
  years = 2021:2024,
  geography = "State",
  geography_values = "56"  
)
file.exists("data/variable_values.rds")
nc_data <- query_pums()

```

```{r}
# Weighted summaries: called automatically by summary(census_data)
summary.census <- function(
    object,
    numeric_vars = NULL,
    categorical_vars = NULL
) {
  is_numeric_variable <- function(x) {
    is.numeric(x) &&
      !inherits(x, c("difftime", "POSIXt", "Date"))
  }

  allowed_numeric <- names(object)[
    vapply(object, is_numeric_variable, logical(1))
  ]
  allowed_numeric <- setdiff(allowed_numeric, c("PWGTP", "year"))

  allowed_categorical <- names(object)[
    vapply(object, function(x) {
      is.factor(x) || is.character(x) || is.logical(x)
    }, logical(1))
  ]
  allowed_categorical <- union(
    allowed_categorical,
    intersect("year", names(object))
  )

  if (is.null(numeric_vars)) {
    numeric_vars <- allowed_numeric
  }

  if (is.null(categorical_vars)) {
    categorical_vars <- allowed_categorical
  }

  if (!is.character(numeric_vars) || anyNA(numeric_vars) ||
      !all(numeric_vars %in% allowed_numeric)) {
    stop("Choose valid numeric variables, excluding PWGTP, year, and times.")
  }

  if (!is.character(categorical_vars) || anyNA(categorical_vars) ||
      !all(categorical_vars %in% allowed_categorical)) {
    stop("Choose valid categorical variables or year.")
  }

  weights <- object$PWGTP

  if (!is.numeric(weights) || any(!is.finite(weights)) ||
      any(weights < 0)) {
    stop("PWGTP must contain finite, nonnegative numeric weights.")
  }

  numeric_results <- lapply(unique(numeric_vars), function(variable) {
    x <- object[[variable]]

    # Exclude missing values and their weights for this variable
    keep <- is.finite(x) & weights > 0
    x <- x[keep]
    w <- weights[keep]

    if (length(x) == 0) {
      return(c(mean = NA_real_, sd = NA_real_))
    }

    weighted_mean <- sum(x * w) / sum(w)

    # Equivalent to the assignment's formula, but more numerically stable
    weighted_sd <- sqrt(
      sum(w * (x - weighted_mean)^2) / sum(w)
    )

    c(mean = weighted_mean, sd = weighted_sd)
  })

  names(numeric_results) <- unique(numeric_vars)

  categorical_results <- lapply(
    unique(categorical_vars),
    function(variable) {
      x <- object[[variable]]

      if (!is.factor(x)) {
        x <- factor(x)
      }

      # Include missing values as a separate category when present
      x <- addNA(x, ifany = TRUE)

      counts <- tapply(weights, x, sum, default = 0)
      setNames(as.numeric(counts), names(counts))
    }
  )

  names(categorical_results) <- unique(categorical_vars)

  list(
    numeric = numeric_results,
    categorical = categorical_results
  )
}


# Weighted boxplots require ggplot2 and quantreg:
# install.packages(c("ggplot2", "quantreg"))

# Called automatically by plot(census_data, ...)
plot.census <- function(x, cat_var, num_var, ...) {
  if (!is.character(cat_var) || length(cat_var) != 1 ||
      is.na(cat_var) || !(cat_var %in% names(x))) {
    stop("Specify one categorical column name as a string.")
  }

  if (!is.character(num_var) || length(num_var) != 1 ||
      is.na(num_var) || !(num_var %in% names(x))) {
    stop("Specify one numeric column name as a string.")
  }

  category <- x[[cat_var]]
  numeric_value <- x[[num_var]]

  if (!(cat_var == "year" || is.factor(category) ||
        is.character(category) || is.logical(category))) {
    stop("cat_var must be categorical or year.")
  }

  if (!is.numeric(numeric_value) ||
      inherits(numeric_value, c("difftime", "POSIXt", "Date")) ||
      num_var %in% c("PWGTP", "year")) {
    stop("num_var must be numeric, excluding PWGTP, year, and times.")
  }

  if (!is.numeric(x$PWGTP) || any(!is.finite(x$PWGTP)) ||
      any(x$PWGTP < 0)) {
    stop("PWGTP must contain finite, nonnegative numeric weights.")
  }

  keep <- !is.na(category) & is.finite(numeric_value) & x$PWGTP > 0
  plot_data <- x[keep, ]

  if (nrow(plot_data) == 0) {
    stop("No complete observations with positive weights to plot.")
  }

  plot_data[[cat_var]] <- factor(plot_data[[cat_var]])

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data[[cat_var]],
      y = .data[[num_var]],
      weight = .data[["PWGTP"]]
    )
  ) +
    ggplot2::geom_boxplot() +
    ggplot2::labs(
      x = cat_var,
      y = num_var,
      title = paste(num_var, "by", cat_var)
    ) +
    ggplot2::theme_minimal()
}


# Summarize all eligible variables
summary(nc_data)

# Summarize selected variables
summary(
  nc_data,
  numeric_vars = "AGEP",
  categorical_vars = "SEX"
)

# Weighted age boxplots by sex
plot(nc_data, cat_var = "SEX", num_var = "AGEP")

# Summarize and plot multiple years
summary(
  combined_data,
  numeric_vars = "AGEP",
  categorical_vars = c("SEX", "year")
)

plot(combined_data, cat_var = "year", num_var = "AGEP")
```
