---
title: "Exploring Census PUMS Data with R"
author: "AMJ"
---

## Introduction

This project is based around developing functions with the Public Use Microdata (PUMS) Census API. The goal of this project is to learn how to work with R in a real-life application and to make something where users would be able to choose variables, survey years, and geographic subsets rather than rewriting the API request each time.

## Setting Things Up

### Packages and API Key

These are all the packages I am using for this project. The `quantreg` and `ggplot2` packages are useful for creating plots!

```{r}
# Read packages
library(httr)
library(jsonlite)
library(tibble)
library(hms)
library(quantreg)
library(ggplot2)
```

#### Load the API Key

The API key is stored in a separate `.Renviron` file in the main project folder, which will allow the API key to be used without it being exposed in the GitHub commits. I did this by making a `.gitignore` file which doesn't push any files to GitHub with the `.Renviron` file type extension. The code below loads the API key.

```{r}
# Import API Key
readRenviron(".Renviron")
stopifnot(nzchar(Sys.getenv("CENSUS_API_KEY")))
```

## Data

### Inspecting the Variable Dictionary

I began with the given URL to understand the JSON structure before writing all the reusable function. So, `GET()` downloads the response, `stop_for_status()` checks for HTTP errors, and `fromJSON()` parses the text into R lists. Setting `simplifyVector = FALSE` preserves the nested list structure.

The following example looks at `SEX`. Its dictionary contains both a description and the mapping between codes and labels.

```{r}
url <- "https://api.census.gov/data/2024/acs/acs1/pums/variables.json"

response <- GET(url)
stop_for_status(response)

# Convert the response to text, then parse it into R lists
json_text <- content(response, as = "text", encoding = "UTF-8")
dictionary <- fromJSON(json_text, simplifyVector = FALSE)

# Inspect one variable
dictionary$variables$SEX
dictionary$variables$SEX$values
```

## Part 1: Data Processing

### Parsing Variable Values

I generalized that process in `get_variable_values(year)`. The function creates the URL for the selected year, keeps the required variables, and extracts each variable's entire `values` element. Doing this preserves both category labels and numeric ranges.

As noted, the state variable is named `ST` in 2021–2022 and `STATE` in 2023–2024, so the function is able to select the name according to the year.

```{r}
get_variable_values <- function(year) {

  # Build the URL for the chosen year
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums/variables.json"
  )

  response <- GET(url)
  stop_for_status(response)

  json_text <- content(response, as = "text", encoding = "UTF-8")
  dictionary <- fromJSON(json_text, simplifyVector = FALSE)

  # State variable name changes after 2022
  state_var <- if (year <= 2022) "ST" else "STATE"

  wanted <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
    "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
    state_var, "REGION", "DIVISION"
  )

  # Keep only the requested variables
  selected <- dictionary$variables[wanted]

  # Extract each variable's values, preserving variable names
  result <- lapply(selected, function(variable) {
    variable$values
  })

  return(result)
}
```

Here, I call the function for each year and save the four named lists in one RDS file which is found in the data folder. All coming functions can read this file to interpret the returned codes using the appropriate year's dictionary.

```{r}
variable_values <- list(
  "2021" = get_variable_values(2021),
  "2022" = get_variable_values(2022),
  "2023" = get_variable_values(2023),
  "2024" = get_variable_values(2024)
)
```

#### Inspect the Selected Values

```{r}
# Check the result
names(variable_values[["2024"]])
variable_values[["2024"]]$SEX
```

#### Save the Dictionaries

```{r}
# Create the folder and save the four lists in one RDS file
dir.create("data", showWarnings = FALSE)
saveRDS(variable_values, file = "data/variable_values.rds")
```

### Parsing API Data

I divided the parsing work into many helper functions. Here is what they are and what they all do:

- `response_to_tibble()` takes a `GET()` response, uses its first row as column names, and converts the remaining rows into a tibble.
- `normalize_codes()` lets codes like `001` and `1` match.
- `time_midpoint()` converts a labeled time interval into its midpoint in seconds, which accounts for any intervals crossing midnight as well as A.M. and P.M.
- `parse_time_codes()` looks up time labels and returns `hms` time-of-day values. Missing or unparseable time intervals become `NA`.
- `parse_category()` converts category codes into labeled factors.

The main `query_pums()` function validates the year, variable selections, geography level, and geography codes before calling the API. It requires `PWGTP`, at least one other numeric variable, and at least one categorical variable. Defaults select age and sex in North Carolina in 2024. Region and Division have separate default codes, and `geography_values = "all"` requests all values within the selected geography level.

After downloading, the function converts columns to their appropriate types and adds the `census` class. This class is what enables the custom summary and plot methods below.

#### Convert an API Response into a Tibble

```{r}
response_to_tibble <- function(response) {
  stop_for_status(response)

  text <- content(response, as = "text", encoding = "UTF-8")

  if (!jsonlite::validate(text)) {
    stop("The API did not return JSON. Check your query and API key.")
  }

  rows <- fromJSON(text, simplifyVector = FALSE)

  if (length(rows) == 0) {
    stop("The API returned no data.")
  }

  headers <- unlist(rows[[1]], use.names = FALSE)

  columns <- lapply(seq_along(headers), function(j) {
    vapply(rows[-1], function(row) {
      if (is.null(row[[j]])) NA_character_ else as.character(row[[j]])
    }, character(1))
  })

  names(columns) <- headers
  as_tibble(columns)
}
```

#### Normalize Variable Codes

```{r}
# Make codes such as "001" and "1" match
normalize_codes <- function(x) {
  x <- trimws(as.character(x))
  numeric_codes <- !is.na(x) & grepl("^[0-9]+$", x)
  x[numeric_codes] <- as.character(as.integer(x[numeric_codes]))
  x
}
```

#### Find the Midpoint of a Time Interval

```{r}
# Convert one time interval label to its midpoint in seconds
time_midpoint <- function(label) {
  if (is.na(label)) {
    return(NA_real_)
  }

  label <- tolower(gsub("\\.", "", label))
  endpoints <- strsplit(label, "\\s+to\\s+")[[1]]

  if (length(endpoints) != 2) {
    return(NA_real_)
  }

  clock_seconds <- function(x) {
    pattern <- "^\\s*([0-9]{1,2}):([0-9]{2})\\s*(am|pm)\\s*$"
    parts <- regmatches(x, regexec(pattern, x))[[1]]

    if (length(parts) != 4) {
      return(NA_real_)
    }

    hour <- as.numeric(parts[2]) %% 12
    minute <- as.numeric(parts[3])

    if (parts[4] == "pm") {
      hour <- hour + 12
    }

    hour * 3600 + minute * 60
  }

  start <- clock_seconds(endpoints[1])
  end <- clock_seconds(endpoints[2])

  if (is.na(start) || is.na(end)) {
    return(NA_real_)
  }

  # Handle intervals that cross midnight
  if (end < start) {
    end <- end + 86400
  }

  ((start + end) / 2) %% 86400
}
```

#### Convert Time Codes into Time Values

```{r}
# Look up time codes and return time-of-day values
parse_time_codes <- function(x, variable_values) {
  labels <- unlist(variable_values$item, use.names = TRUE)
  midpoints <- vapply(labels, time_midpoint, numeric(1))

  index <- match(
    normalize_codes(x),
    normalize_codes(names(labels))
  )

  hms::as_hms(unname(midpoints[index]))
}
```

#### Convert Categorical Codes into Factors

```{r}
# Convert categorical codes into labeled factors
parse_category <- function(x, variable_values) {
  labels <- unlist(variable_values$item, use.names = TRUE)
  codes <- normalize_codes(names(labels))

  # Sort levels by their underlying numeric codes
  ordering <- order(suppressWarnings(as.numeric(codes)), codes)
  labels <- labels[ordering]
  codes <- codes[ordering]

  factor(
    normalize_codes(x),
    levels = codes,
    labels = unname(labels)
  )
}
```

#### Query and Clean One Survey Year

```{r}
query_pums <- function(
    year = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_values = NULL,
    rds_file = "data/variable_values.rds",
    api_key = Sys.getenv("CENSUS_API_KEY", unset = "")
) {
  # Validate year
  if (length(year) != 1 || is.na(year) ||
      !(year %in% 2021:2024)) {
    stop("year must be one of 2021, 2022, 2023, or 2024.")
  }

  allowed_numeric <- c(
    "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"
  )

  allowed_categorical <- c("FER", "HHL", "SCH", "SCHL", "SEX")

  # Validate numeric variables
  if (!is.character(numeric_vars) || anyNA(numeric_vars) ||
      !all(numeric_vars %in% allowed_numeric)) {
    stop("Invalid numeric variable.")
  }

  if (!("PWGTP" %in% numeric_vars) ||
      length(setdiff(numeric_vars, "PWGTP")) == 0) {
    stop("Include PWGTP and at least one other numeric variable.")
  }

  # Validate categorical variables
  if (!is.character(categorical_vars) ||
      length(categorical_vars) == 0 ||
      anyNA(categorical_vars) ||
      !all(categorical_vars %in% allowed_categorical)) {
    stop("Include at least one valid categorical variable.")
  }

  # Validate geography level
  if (!is.character(geography) || length(geography) != 1 ||
      is.na(geography) ||
      !(tolower(geography) %in% c("region", "division", "state"))) {
    stop("geography must be exactly one of Region, Division, or State.")
  }

  geography <- tolower(geography)
  numeric_vars <- unique(numeric_vars)
  categorical_vars <- unique(categorical_vars)

  # Load the dictionaries saved in Part 1
  dictionaries <- readRDS(rds_file)
  metadata <- dictionaries[[as.character(year)]]

  if (is.null(metadata)) {
    stop("The RDS file does not contain the requested year.")
  }

  geography_variable <- switch(
    geography,
    state = if (year <= 2022) "ST" else "STATE",
    region = "REGION",
    division = "DIVISION"
  )

  geography_labels <- unlist(
    metadata[[geography_variable]]$item,
    use.names = TRUE
  )

  valid_codes <- normalize_codes(names(geography_labels))

  # Defaults: North Carolina, South Atlantic, or South
  if (is.null(geography_values)) {
    geography_values <- switch(
      geography,
      state = "37",
      division = "5",
      region = "3"
    )
  }

  if (length(geography_values) == 0 || anyNA(geography_values)) {
    stop("Provide valid geography codes or 'all'.")
  }

  geography_values <- as.character(geography_values)

  if (any(tolower(geography_values) == "all")) {
    if (length(geography_values) != 1) {
      stop("'all' cannot be combined with specific geography codes.")
    }

    geography_filter <- "*"
  } else {
    codes <- normalize_codes(geography_values)

    if (!all(codes %in% valid_codes)) {
      stop("Invalid geography code. Check the selected year's RDS values.")
    }

    # State FIPS codes must have two digits
    if (geography == "state") {
      codes <- sprintf("%02d", as.integer(codes))
    }

    geography_filter <- paste(unique(codes), collapse = ",")
  }

  # Build and submit the API query
  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums"
  )

  query <- list(
    get = paste(c(numeric_vars, categorical_vars), collapse = ",")
  )

  query[["for"]] <- paste0(geography, ":", geography_filter)

  if (nzchar(api_key)) {
    query$key <- api_key
  }

  response <- GET(url, query = query)
  census_data <- response_to_tibble(response)

  # Convert numeric and time variables
  for (variable in numeric_vars) {
    if (variable %in% c("JWAP", "JWDP")) {
      census_data[[variable]] <- parse_time_codes(
        census_data[[variable]],
        metadata[[variable]]
      )
    } else {
      census_data[[variable]] <- suppressWarnings(
        as.numeric(census_data[[variable]])
      )
    }
  }

  # Convert categorical variables using the correct year's labels
  for (variable in categorical_vars) {
    census_data[[variable]] <- parse_category(
      census_data[[variable]],
      metadata[[variable]]
    )
  }

  class(census_data) <- c("census", class(census_data))
  return(census_data)
}
```

### Combining Years

`query_pums_years()` calls the single-year function for each distinct requested year. Each result receives a `year` column, and `bind_rows()` combines the results into one census tibble. Assigning the returned value to an object such as `combined_data` makes it available to later chunks.

```{r}
query_pums_years <- function(
    years = 2024,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = "SEX",
    geography = "State",
    geography_values = NULL,
    rds_file = "data/variable_values.rds",
    api_key = Sys.getenv("CENSUS_API_KEY", unset = "")
) {
  if (length(years) == 0 || anyNA(years) ||
      !all(years %in% 2021:2024)) {
    stop("years must contain only 2021, 2022, 2023, or 2024.")
  }

  yearly_data <- lapply(unique(years), function(year) {
    data <- query_pums(
      year = year,
      numeric_vars = numeric_vars,
      categorical_vars = categorical_vars,
      geography = geography,
      geography_values = geography_values,
      rds_file = rds_file,
      api_key = api_key
    )

    data$year <- as.integer(year)
    data
  })

  combined_data <- dplyr::bind_rows(yearly_data)

  class(combined_data) <- unique(c("census", class(combined_data)))
  return(combined_data)
}
```

## Part 2: Writing a Generic Function for Summarizing

### Generic Functions

R uses the `census` class to run our custom `summary.census()` and `plot.census()` methods when we call `summary()` or `plot()`.

The summary function returns a named list containing:

- Weighted means and standard deviations for numeric variables.
- Weighted counts for categorical variables.

By default, it summarizes all eligible columns. It excludes `PWGTP` and time variables from numeric summaries and treats `year` as categorical. Users can choose specific columns or use `character(0)` to skip either type of summary.

Here, $x_i$ is a numeric value and $w_i$ is its `PWGTP` weight. Missing numeric values and their weights are excluded from that calculation. Missing categorical values are counted separately.

The plotting function creates a boxplot for one numeric variable across the groups of one categorical variable. It includes `weight = PWGTP` to account for each row's weight. Time variables and `PWGTP` cannot be used as the numeric plotting variable.

```{r}
# Weighted summaries: called automatically by summary(census_data)
summary.census <- function(
    object,
    numeric_vars = NULL,
    categorical_vars = NULL
) {
  # Find numeric columns, excluding time variables, weights, and year
  numeric_columns <- names(object)[
    vapply(object, function(x) {
      is.numeric(x) &&
        !inherits(x, c("difftime", "POSIXt", "Date"))
    }, logical(1))
  ]

  numeric_columns <- setdiff(numeric_columns, c("PWGTP", "year"))

  # Find categorical columns, including year
  categorical_columns <- names(object)[
    vapply(object, function(x) {
      is.factor(x) || is.character(x) || is.logical(x)
    }, logical(1))
  ]

  categorical_columns <- union(
    categorical_columns,
    intersect("year", names(object))
  )

  # Default to all eligible columns
  if (is.null(numeric_vars)) {
    numeric_vars <- numeric_columns
  }

  if (is.null(categorical_vars)) {
    categorical_vars <- categorical_columns
  }

  # Check the selected variables
  if (!is.character(numeric_vars) || anyNA(numeric_vars) ||
      !all(numeric_vars %in% numeric_columns)) {
    stop("Choose valid numeric variables, excluding PWGTP, year, and times.")
  }

  if (!is.character(categorical_vars) || anyNA(categorical_vars) ||
      !all(categorical_vars %in% categorical_columns)) {
    stop("Choose valid categorical variables or year.")
  }

  weights <- object$PWGTP

  if (!is.numeric(weights) || any(!is.finite(weights)) ||
      any(weights < 0)) {
    stop("PWGTP must contain finite, nonnegative numeric weights.")
  }

  # Calculate weighted means and standard deviations
  numeric_results <- list()

  for (variable in numeric_vars) {
    numeric_vector <- object[[variable]]
    weight_vector <- weights

    # Remove missing values and their corresponding weights
    keep <- is.finite(numeric_vector) & weight_vector > 0
    numeric_vector <- numeric_vector[keep]
    weight_vector <- weight_vector[keep]

    if (length(numeric_vector) == 0) {
      sample_mean <- NA_real_
      sample_sd <- NA_real_
    } else {
      sample_mean <- sum(numeric_vector * weight_vector) /
        sum(weight_vector)

      sample_sd <- sqrt(
        sum(numeric_vector^2 * weight_vector) /
          sum(weight_vector) - sample_mean^2
      )
    }

    numeric_results[[variable]] <- c(
      mean = sample_mean,
      sd = sample_sd
    )
  }

  # Calculate weighted counts for categorical variables
  categorical_results <- list()

  for (variable in categorical_vars) {
    category <- object[[variable]]

    if (!is.factor(category)) {
      category <- factor(category)
    }

    category <- addNA(category, ifany = TRUE)

    counts <- tapply(weights, category, sum, default = 0)

    categorical_results[[variable]] <- setNames(
      as.numeric(counts),
      names(counts)
    )
  }

  # Return a named list
  return(list(
    numeric = numeric_results,
    categorical = categorical_results
  ))
}
```

### Graphical Summaries

This method creates a weighted boxplot for the chosen numeric and categorical columns.

```{r}
# Called automatically by plot(census_data, ...)
plot.census <- function(x, cat_var, num_var, ...) {
  if (!is.character(cat_var) || length(cat_var) != 1 ||
      is.na(cat_var) || !(cat_var %in% names(x))) {
    stop("Specify one categorical column name as a string.")
  }

  if (!is.character(num_var) || length(num_var) != 1 ||
      is.na(num_var) || !(num_var %in% names(x))) {
    stop("Specify one numeric column name as a string.")
  }

  category <- x[[cat_var]]
  numeric_value <- x[[num_var]]

  if (!(cat_var == "year" || is.factor(category) ||
        is.character(category) || is.logical(category))) {
    stop("cat_var must be categorical or year.")
  }

  if (!is.numeric(numeric_value) ||
      inherits(numeric_value, c("difftime", "POSIXt", "Date")) ||
      num_var %in% c("PWGTP", "year")) {
    stop("num_var must be numeric, excluding PWGTP, year, and times.")
  }

  if (!is.numeric(x$PWGTP) || any(!is.finite(x$PWGTP)) ||
      any(x$PWGTP < 0)) {
    stop("PWGTP must contain finite, nonnegative numeric weights.")
  }

  keep <- !is.na(category) & is.finite(numeric_value) & x$PWGTP > 0
  plot_data <- x[keep, ]

  if (nrow(plot_data) == 0) {
    stop("No complete observations with positive weights to plot.")
  }

  plot_data[[cat_var]] <- factor(plot_data[[cat_var]])

  ggplot2::ggplot(
    plot_data,
    ggplot2::aes(
      x = .data[[cat_var]],
      y = .data[[num_var]],
      weight = .data[["PWGTP"]]
    )
  ) +
    ggplot2::geom_boxplot() +
    ggplot2::labs(
      x = cat_var,
      y = num_var,
      title = paste(num_var, "by", cat_var)
    ) +
    ggplot2::theme_minimal()
}
```

## Testing the Functions

### Dictionary structure and time parsing

These checks verify that all four dictionaries contain the required variables and that time conversion handles morning, noon, midnight, and missing values. A failed `stopifnot()` stops rendering, so a success message appears only if its checks pass.

```{r}
for (y in 2021:2024) {
  state_name <- if (y <= 2022) "ST" else "STATE"
  expected <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP",
                "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX",
                state_name, "REGION", "DIVISION")
  d <- variable_values[[as.character(y)]]
  stopifnot(setequal(names(d), expected),
            all(vapply(d, function(x) !is.null(x), logical(1))))
}
```

#### Test Time Midpoints

```{r}
stopifnot(
  time_midpoint("6:00 a.m. to 6:04 a.m.") == 21720,
  time_midpoint("12:00 p.m. to 12:04 p.m.") == 43320,
  time_midpoint("12:00 a.m. to 12:04 a.m.") == 120,
  is.na(time_midpoint("N/A (not a worker)"))
)
cat("Dictionary and time parsing checks passed.")
```

### Checking weighted calculations by hand

For ages 20 and 40 with weights 1 and 3, the weighted mean is 35 and the weighted standard deviation is $\sqrt{75}$, or approximately 8.66. The weighted counts should be 1 for group A and 3 for group B. This small example checks the arithmetic independently of the API.

```{r}
toy_data <- tibble::tibble(
  AGEP = c(20, 40), PWGTP = c(1, 3), group = factor(c("A", "B"))
)
class(toy_data) <- c("census", class(toy_data))
toy_summary <- summary(toy_data)
stopifnot(
  isTRUE(all.equal(unname(toy_summary$numeric$AGEP["mean"]), 35)),
  isTRUE(all.equal(unname(toy_summary$numeric$AGEP["sd"]), sqrt(75))),
  toy_summary$categorical$group["A"] == 1,
  toy_summary$categorical$group["B"] == 3
)
toy_summary
cat("Weighted calculation checks passed.")
```

### Single-year and multiple-year downloads

I test the defaults with North Carolina and the multiple-year function with Wyoming (state code `56`). Wyoming keeps the four-year test relatively small. These are separate datasets: `nc_data` describes North Carolina in 2024; `combined_data` describes Wyoming in 2021–2024.

```{r}
combined_data <- query_pums_years(
  years = 2021:2024,
  geography = "State",
  geography_values = "56"  
)
```

#### Download One Year

```{r}
file.exists("data/variable_values.rds")
nc_data <- query_pums()
```

```{r}
stopifnot(
  nrow(nc_data) > 0,
  inherits(nc_data, "census"),
  is.numeric(nc_data$AGEP), is.numeric(nc_data$PWGTP),
  is.factor(nc_data$SEX),
  inherits(combined_data, "census"),
  setequal(unique(combined_data$year), 2021:2024)
)
knitr::kable(as.data.frame(table(combined_data$year)),
             col.names = c("Survey year", "Downloaded records"))
cat("Single-year and multiple-year checks passed.")
```

### Summary and plot methods with different inputs

First, I call the default summary and then select age and sex explicitly. I also select `year` in the combined data. The plots use different categorical inputs: sex for North Carolina and year for Wyoming. These weighted age plots demonstrate method dispatch and provide a visual check of the returned data.

The combined summary pools the four annual samples. Its weighted counts across years should not be interpreted as a count of unique people.

```{r}
# Summarize all eligible variables
summary(nc_data)
```

#### Summarize Selected Columns

```{r}
# Summarize selected variables
summary(
  nc_data,
  numeric_vars = "AGEP",
  categorical_vars = "SEX"
)
```

#### Plot Age by Sex

```{r}
# Weighted age boxplots by sex
plot(nc_data, cat_var = "SEX", num_var = "AGEP")+ggplot2::theme_gray()
```

#### Summarize Multiple Years

```{r}
# Summarize and plot multiple years
summary(
  combined_data,
  numeric_vars = "AGEP",
  categorical_vars = c("SEX", "year")
)
```

#### Plot Age by Year

```{r}
plot(combined_data, cat_var = "year", num_var = "AGEP")+ggplot2::theme_gray()
```

### Invalid inputs

The following tests deliberately supply invalid arguments. Each test checks that an error occurs and displays its message while allowing the document to continue. This verifies that the functions reject invalid requests rather than silently downloading unintended data.

```{r}
expect_error <- function(expr) {
  message <- tryCatch({ force(expr); NULL },
                      error = function(e) conditionMessage(e))
  if (is.null(message)) stop("Test failed: an error was expected.")
  message
}
```

#### Check Invalid Requests

```{r}
error_checks <- list(
  "Invalid year" = expect_error(query_pums(year = 2020)),
  "Multiple years in single-year query" = expect_error(query_pums(year = c(2023, 2024))),
  "Missing weight" = expect_error(query_pums(numeric_vars = "AGEP")),
  "Only weight" = expect_error(query_pums(numeric_vars = "PWGTP")),
  "Invalid numeric variable" = expect_error(query_pums(numeric_vars = c("PWGTP", "BAD"))),
  "Missing category" = expect_error(query_pums(categorical_vars = character(0))),
  "Invalid category" = expect_error(query_pums(categorical_vars = "BAD")),
  "Multiple geography levels" = expect_error(query_pums(geography = c("State", "Region"))),
  "Invalid geography level" = expect_error(query_pums(geography = "All")),
  "Invalid state code" = expect_error(query_pums(geography_values = "99")),
  "Invalid plot input" = expect_error(plot(nc_data, cat_var = "AGEP", num_var = "SEX"))
)
knitr::kable(data.frame(Test = names(error_checks),
                        Message = unlist(error_checks, use.names = FALSE)))
```

These examples test the State geography option. They do not constitute a live test of every Region, Division, or all-geographies query.

## Investigation

### Commuting Time in North Carolina

#### Question and Data Selection

Among North Carolina adults with a positive reported commute in 2024, how does commuting time differ by age group and sex?

I use `JWMNP` for travel time to work in minutes, `AGEP` for age, and `SEX` for the sex categories returned by the Census. I also request `JWDP` to verify time-column conversion. The query below demonstrates different numeric inputs from the earlier age-only examples.

```{r}
commute_data <- query_pums(
  year = 2024,
  numeric_vars = c("AGEP", "PWGTP", "JWMNP", "JWDP"),
  categorical_vars = "SEX",
  geography = "State",
  geography_values = "37"
)
stopifnot(is.numeric(commute_data$JWMNP),
          inherits(commute_data$JWDP, "hms"))
head(commute_data[c("AGEP", "JWMNP", "JWDP", "SEX")])
```

#### Select Adult Commuters

```{r}
# Keep adults with valid, positive travel time and positive weight.
keep <- with(commute_data,
             is.finite(AGEP) & AGEP >= 18 &
             is.finite(JWMNP) & JWMNP > 0 &
             is.finite(PWGTP) & PWGTP > 0 & !is.na(SEX))
commuters <- commute_data[keep, ]
stopifnot(nrow(commuters) > 0)
```

#### Create Age Groups

```{r}
commuters$age_group <- cut(
  commuters$AGEP,
  breaks = c(18, 30, 45, 65, Inf), right = FALSE,
  labels = c("18–29", "30–44", "45–64", "65+")
)
knitr::kable(data.frame(
  Stage = c("All downloaded records", "Included adult commuters"),
  Records = c(nrow(commute_data), nrow(commuters))
))
```

This restriction avoids treating not-applicable or zero commute values as actual travel times. It excludes people without a positive reported commute, including people working from home, so the results describe this selected group rather than all adults.

### Overall weighted summary

Here, I summarize two numeric variables and two categorical variables in one call. The time column `JWDP` is not included in the numeric summary.

```{r}
commute_summary <- summary(
  commuters,
  numeric_vars = c("AGEP", "JWMNP"),
  categorical_vars = c("SEX", "age_group")
)
commute_summary
overall_minutes <- unname(commute_summary$numeric$JWMNP["mean"])
overall_sd <- unname(commute_summary$numeric$JWMNP["sd"])
```

The weighted mean commute in the included sample is **`r round(overall_minutes, 2)` minutes**, with a weighted standard deviation of **`r round(overall_sd, 2)` minutes**. The categorical counts are sums of person weights, not the number of downloaded rows.

### Comparison by age group

I use the custom plot method with a newly created categorical variable, `age_group`, and a different numeric variable, `JWMNP`. The boxplot shows distributions, while the table reports group means and standard deviations.

```{r}
plot(commuters, cat_var = "age_group", num_var = "JWMNP") +
  ggplot2::labs(title = "Commuting time by age group: North Carolina, 2024",
                x = "Age group", y = "Travel time to work (minutes)")+ggplot2::theme_gray()
```

#### Summarize Each Group

The helper below calls `summary()` separately for each group and collects the results in a table.

```{r}
# This helper uses summary.census() for every group's calculations.
group_summary_table <- function(data, group_var) {
  groups <- levels(droplevels(data[[group_var]]))
  tables <- lapply(groups, function(g) {
    subset <- data[!is.na(data[[group_var]]) & data[[group_var]] == g, ]
    result <- summary(subset, numeric_vars = "JWMNP",
                      categorical_vars = group_var)
    data.frame(
      Group = g,
      Weighted_count = sum(result$categorical[[group_var]]),
      Mean_minutes = unname(result$numeric$JWMNP["mean"]),
      SD_minutes = unname(result$numeric$JWMNP["sd"])
    )
  })
  do.call(rbind, tables)
}
```

#### Report the Age-Group Results

```{r}
age_results <- group_summary_table(commuters, "age_group")
knitr::kable(age_results, digits = 2,
             col.names = c("Age group", "Weighted count", "Mean minutes", "SD minutes"))
age_high <- age_results$Group[which.max(age_results$Mean_minutes)]
age_low <- age_results$Group[which.min(age_results$Mean_minutes)]
age_gap <- diff(range(age_results$Mean_minutes))
```

The **`r age_high`** group has the highest weighted mean commute and the **`r age_low`** group has the lowest. The difference between the largest and smallest group means is **`r round(age_gap, 2)` minutes**. If groups tie, this identifies the first tied group in table order. The boxplot complements these averages by showing the spread within each group.

### Comparison by sex

I next change the plot's categorical input to `SEX` and calculate separate weighted summaries for its levels. The grouping table calls my summary method on each subset; the summary method itself does not automatically group numeric statistics by the requested categorical variables.

```{r}
plot(commuters, cat_var = "SEX", num_var = "JWMNP") +
  ggplot2::labs(title = "Commuting time by sex: North Carolina, 2024",
                x = "Sex", y = "Travel time to work (minutes)")+ggplot2::theme_gray()
```

#### Report the Sex-Group Results

```{r}
sex_results <- group_summary_table(commuters, "SEX")
knitr::kable(sex_results, digits = 2,
             col.names = c("Sex", "Weighted count", "Mean minutes", "SD minutes"))
sex_high <- sex_results$Group[which.max(sex_results$Mean_minutes)]
sex_gap <- diff(range(sex_results$Mean_minutes))
```

The largest weighted mean in this comparison is for **`r sex_high`**, and the range of the group means is **`r round(sex_gap, 2)` minutes**. These are descriptive differences; they do not establish that sex causes differences in commuting time.

## Conclusion

For the North Carolina investigation, the overall weighted mean is **`r round(overall_minutes, 2)` minutes**. The age-group mean range is **`r round(age_gap, 2)` minutes**, and the sex-group mean range is **`r round(sex_gap, 2)` minutes**.
